---
title: "Multivariate Normal Distribution"
author: "Pulong Ma"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(root.dir="~/Documents/Teaching/STAT475_575/Fall2024/Data/")

library(ggplot2)
#setwd("~/Documents/Teaching/STAT475_575/Fall2024/Data/")
```

# Stiffness of boards

```{r}
# load the data
setwd("~/Documents/Teaching/STAT475_575/Fall2024/Data/")
boards <-read.table(file="./board.stiffness.dat",header=F)

boards.var <- c('board','x1','x2','x3','x4')
names(boards) <- boards.var
```

## Compute sample correlation/covariance matrix
```{r}
bcorr <- cor(boards[ ,2:5])
print(bcorr)

bvar <- var(boards[ ,2:5])
print(bvar)
```

## Create a scatterplot matrix
```{r}
pairs(boards[ ,2:5])
```

## Create a normal probability plot
```{r}
par(mfrow=c(2,2),pch=5)
for (i in 2:ncol(boards)){
  qqnorm(boards[,i], main="Normal Q-Q Plot") 
}
```

## Compute the Shapiro-Wilk test statistic for each variable
```{r}
apply(boards[ ,2:5],2,shapiro.test)
```


# Test multivariate normality

## Compute Shapiro-Wilk test for multivariate normality
```{r}
# install.packages("mvShapiroTest")
library(mvShapiroTest)
mvShapiro.Test(as.matrix(boards[ , 2:5]))
```

## Generate random samples from a multivariate normal distribution
```{r}
set.seed(4750)
library(MASS)
xsigma <- matrix(c(1, -0.5, -0.5, 1),  ncol=2,byrow=T)
sim.mvn <- data.frame(mvrnorm(n=100, mu=c(0,0), Sigma=xsigma))
head(sim.mvn)
mvShapiro.Test(as.matrix(sim.mvn))
```


# Compute eigenvectors and eigenvalues of the sample covariance matrix
```{r}
evbar = eigen(bvar)
print(evbar)
```

# Outlier detection
```{r}
## Finding mahalanobis distance
xbar = colMeans(boards[,2:5])
dsq = mahalanobis(x=boards[,2:5], center=xbar, cov=bvar)

## Cutoff value for distances from a chisquare distribution
cutoff = qchisq(p=0.95, df=nrow(bvar)) # 95th percentile 

## plot observations whose distance is greater than cutoff value
boards[dsq>cutoff,]

```