---
title: "Factor Analysis"
author: "Pulong Ma"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir="~/Documents/Teaching/STAT475_575/Fall2024/Data/")

library(ggplot2)
rm(list=ls())
```


# Motivating example

In this example, we'll use six different personality traits, each assessed by a few items/questions. Factor analysis will help us reduce these traits into a smaller number of underlying factors, which can explain the relationships between the items.

Let’s consider a personality questionnaire designed to measure six traits:

- Openness
- Conscientiousness
- Extraversion
- Agreeableness
- Neuroticism
- Emotional Stability

Each trait is measured by several items (questions). Our goal is to use factor analysis to see if these items cluster together in ways that reveal underlying factors.


```{r}
# Load required packages
#install.packages(c("psych", "ggplot2", "corrplot", "GPArotation"))
library(psych)
library(ggplot2)
library(corrplot)
library(GPArotation)

# Generate mock data for 6 personality traits and 100 individuals
set.seed(123)
set.seed(100)
data <- data.frame(
  Openness_1 = rnorm(200, mean=4, sd=0.7),  # Openness
  Openness_2 = rnorm(200, mean=3.5, sd=0.8),
  Openness_3 = rnorm(200, mean=4, sd=0.7),
  Conscientiousness_1 = rnorm(200, mean=3.8, sd=0.6),  # Conscientiousness
  Conscientiousness_2 = rnorm(200, mean=4, sd=0.5),
  Conscientiousness_3 = rnorm(200, mean=3.9, sd=0.6),
  Extraversion_1 = rnorm(200, mean=3.2, sd=0.9),  # Extraversion
  Extraversion_2 = rnorm(200, mean=3.5, sd=0.8),
  Extraversion_3 = rnorm(200, mean=3, sd=0.9),
  Agreeableness_1 = rnorm(200, mean=4.2, sd=0.5),  # Agreeableness
  Agreeableness_2 = rnorm(200, mean=4, sd=0.6),
  Agreeableness_3 = rnorm(200, mean=4.1, sd=0.5),
  Neuroticism_1 = rnorm(200, mean=2.8, sd=1),  # Neuroticism
  Neuroticism_2 = rnorm(200, mean=3, sd=0.9),
  Neuroticism_3 = rnorm(200, mean=2.9, sd=1),
  Emotional_Stability_1 = rnorm(200, mean=3.5, sd=0.7),  # Emotional Stability
  Emotional_Stability_2 = rnorm(200, mean=3.6, sd=0.6),
  Emotional_Stability_3 = rnorm(200, mean=3.7, sd=0.5)
)


# KMO measure
kmo_result <- KMO(cor(data))
kmo_result

# Bartlett's test
bartlett_result <- cortest.bartlett(cor(data), n = nrow(data))
bartlett_result

```

```{r}
# Perform factor analysis
factor_result <- factanal(data, factors = 6, rotation = "varimax", scores="regression")
print(factor_result, digits = 2, cutoff = 0.3)

```


```{r}
# Plot factor loadings
loadings <- factor_result$loadings[,1:6]
corrplot(loadings, is.corr = FALSE, method = "circle", tl.cex = .9)

```

```{r}
# Create a scree plot
eigenvalues <- eigen(cor(data))$values
scree_data <- data.frame(
  Factor = 1:length(eigenvalues),
  Eigenvalue = eigenvalues
)

ggplot(scree_data, aes(x = Factor, y = Eigenvalue)) +
  geom_point() +
  geom_line() +
  ggtitle("Scree Plot") +
  xlab("Factor") + 
  ylab("Eigenvalue")

```

```{r}
# Extract factor scores
factor_scores <- as.data.frame(factor_result$scores)
names(factor_scores) <- c("Factor1", "Factor2")

# Scatter plot of factors
ggplot(factor_scores, aes(x = Factor1, y = Factor2)) +
  geom_point() +
  ggtitle("Factor 1 vs Factor 2") +
  xlab("Factor 1") + 
  ylab("Factor 2")

```

# Life expectance data
```{r cars}
setwd("~/Documents/Teaching/STAT475_575/Fall2024/Data/")
life<-read.csv("life.csv",  row.names=1)
head(life)

```

## Examine the correlation matrix and scatterplot matrix
```{r}
lifecor <- cor(life)
lifecor
```
#  Construct a scatterplot of the data
```{r}
pairs(life)
```

## Compute PCs and factor loadings

```{r}
lifepc <- prcomp(life, scale=T, center=T)
summary(lifepc)
```

# produce a scree plot

```{r}
plot(lifepc$sdev^2, xlab="Component Number",
        ylab="Component Variance (eigenvalue)",  
        main="Scree Diagram", type="l")
```

# rotations/PCs
```{r}
lifepc$rotation
```


## Compute factor loadings for the three factors using PCs from the sample correlation matrix

```{r}
lifepc$loadings <- lifepc$rotation%*%diag(lifepc$sdev)
lifepc$loadings
```


## Varimax rotation of loadings for three factors computed from PCs


```{r}
varimax(lifepc$loadings[ ,1:2])
```


## MLE 

```{r}
# Compute p-values for tests for numbers of factors
sapply(1:3, function(f)
    factanal(life, factors = f, method ="mle")$PVAL)

# Compute loadings for a three factor model
# with no rotation.  Factor scores are not computed.
factanal(life, factors = 3, method ="mle", rotation="none")
```

## Factor scores

Recall that we have the following orthogonal factor model 
$$
\mathbf{X}_i = \mu + L F_i + \epsilon_i, i=1,\ldots, n,
$$
where $\mathbf{X}_i$ is the $i$-the observation with $p$ variables, $L$ is an $p\times m$ matrix, $F_i$ is the $m$-by-1 vector of factor scores for $i$-th observation. Each entry in $F_i$ corresponds to one of the common factors. 

Once the factor loading matrix $L$ is found using the methods introduced in class, we then need to find factor scores. 

There are three common ways to estimate factor scores:
- Ordinary Least Squares (OLS)
- Weighted Least Squares (WLS)
- Regression method


# Compute factor scores with varimax rotation 
```{r}
# Compute variamx rotation (default)of factor loadings
#  Factor scores are also computed using the regression (default)
#  method
life_varimax <- factanal(life, factors = 3, method ="mle",
                    rotation="varimax", scores="regression")
life_varimax

# Print all of the loading values
print(loadings(life_varimax), cutoff = 1e-05)

names(life_varimax)
life_varimax$scores

```



# Plot the scores
```{r}
scores <- life_varimax$scores
plot(scores[,1], scores[,2], type = "n", xlab = "Factor 1",
        ylab = "Factor 2")
 text(scores[,1], scores[,2], abbreviate(rownames(life), 6),
      cex = 0.8)
 plot(scores[,1], scores[,3], type = "n", xlab = "Factor 1",
         ylab = "Factor 3")
 text(scores[,1], scores[,3], abbreviate(rownames(life), 6),
       cex = 0.8)
 plot(scores[,2], scores[,3], type = "n", xlab = "Factor 2",
        ylab = "Factor 3")
 text(scores[,2], scores[,3], abbreviate(rownames(life), 6),
      cex = 0.8)

```

# Perform factor analysis with 2 factors
```{r}
life_varimax2 <- factanal(life, factors = 2, method ="mle",
                    rotation="varimax", scores="regression")
life_varimax2
```


# Measure of sampling adequacy
```{r}
library(psych)
KMO(life)
alpha(life)
```

# Factor rotaton

Why do we need to perform rotation?
- Because we wish to find a rotation such that each of the $p$ variables has a **high loading on only one factor**  

Rotation methods: 
- Varimax
- Quartimax rotation
- PROMAX 

```{r}
library(GPArotation)
life_varimax <- factanal(life, factors = 3, method ="mle",
                   rotation="varimax", scores="regression")
print(life_varimax)
quartimax(life_varimax$loadings)
promax(life_varimax$loadings)

# Compute the matrix of correlations between factors.
 M <- promax(life_varimax$loadings, m=2)$rotmat
corfac <- solve(t(M) %*% M)
corfac
```

# Summary

- How do we interpret factor loadings
- How to estimate factor loadings?
  - Principle component method
  - Principle factor method
  - Maximum likelihood method
- How communalities can be used to assess the adequacy of a factor model?
- Likelihood ratio test 
- Factor rotation
- Methods for estimating factor scores


