---
title: "Inference for two or more population means"
author: "Pulong Ma"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(root.dir="~/Documents/Teaching/STAT475_575/Fall2024/Data/")

library(ggplot2)
```


# Two Processes for Manufacturing Soap

```{r}
setwd("~/Documents/Teaching/STAT475_575/Fall2024/Data/")
steel <- read.table(file = "./steel.dat",
          header=F, col.names=c("temp", "yield", "strength"))
steel
```

## Summary statistics for each treatment
```{r}
#  Compute sample mean vector and sample
xbar1<-sapply(steel[steel$temp==1, -1], mean)
xbar1
   
#  covariance matrix for each temperature
xvar1<-var(steel[steel$temp==1 , -1])
xvar1

#  Compute sample mean vector and sample
xbar2<-sapply(steel[steel$temp==2, -1], mean)
xbar2
   
#  covariance matrix for each temperature
xvar2<-var(steel[steel$temp==2 , -1])
xvar2
```

## Scatterplot
```{r}
plot(steel[ ,2], steel[ ,3],
          xlab = "Yield Point",
          ylab = "Ultimate Strength", 
          pch=c(1,16)[steel$temp], cex=1.5)
    abline(lm(steel[steel$temp==1 ,3] ~ 
		             steel[steel$temp==1 ,2]), lty=4, lwd=2)
    abline(lm(steel[steel$temp==2 ,3] ~ 
		             steel[steel$temp==2 ,2]), lty=1, lwd=2)
    legend("bottomright", c("Low Temp", "High Temp"), 
       lty = c(4, 1), lwd=2, pch = c(1, 16), merge = TRUE)

```

## Test for univariate normality 
```{r}
# for the first rolling temp
apply(steel[steel$temp == 1, -1], 2, shapiro.test)

# for the second rolling temp
apply(steel[steel$temp == 2, -1], 2, shapiro.test)
```

## Test for bivariate normality
```{r}
library(mvShapiroTest) 
mvShapiro.Test(as.matrix(steel[ , 2:3]))
```

## Test homogeneity of covariance matrices
```{r}
#  Apply Box's M-test to test the null hypothesis of 
#  homogeneous covariance matrices.  This test is sensitive 
#  to the lack of normality and has a tendency to reject
#  the null hypothesis of homogeneous covariance matrices 
#  just due to lack of normality.

library(biotools)
boxM(steel[ , -1], steel$temp)

```

## Two-sample Hotelling T-square test
```{r}
library(DescTools)

T2result <- HotellingsT2Test(steel[steel$temp == 1, -1], 
             steel[steel$temp == 2, -1])

T2result
```

## Construct T-sqaure confidence intervals
```{r}
#  Compute sample sizes and number of variables
n1 <- dim(steel[steel$temp==1, -1])[1]
n2 <- dim(steel[steel$temp==2, -1])[1]
p <- dim(steel[steel$temp==2, -1])[2]

#  Enter the confidence level
level <- 0.95

# Compute degrees of freedom and the multipliers
df1 <- p
df2 <- n1+n2-p-1
df3 <- n1+n2-2

c_T2 <-  sqrt((n1+n2-2)*p*qf(level,df1,df2)/(n1+n2-p-1))
level2 <- 1-(1-level)/2
c_bon  <- qt(level2, df3)

# Compute pooled covariance matrix 
vpool <- ((n1-1)*xvar1 +(n2-1)*xvar2)/(n1+n2-2)

#  Compute T-squared limits
lower_limit <- (xbar1-xbar2) - c_T2*sqrt(diag(vpool)*((1/n1)+(1/n2)))
upper_limit <- (xbar1-xbar2) + c_T2*sqrt(diag(vpool)*((1/n1)+(1/n2)))

rbind(lower_limit, upper_limit)

```


## Compute Bonferroni limits
```{r}
lower_limit <- (xbar1-xbar2) - c_bon*sqrt(diag(vpool)*((1/n1)+(1/n2)))
upper_limit <- (xbar1-xbar2) + c_bon*sqrt(diag(vpool)*((1/n1)+(1/n2)))
 
rbind(lower_limit, upper_limit)

```

# MANOVA Example

```{r}
#  First enter the data into a data frame.
setwd("~/Documents/Teaching/STAT475_575/Fall2024/Data/")
morel<-read.table("./morel.dat",
                  header=F, col.names=c("studentgroup", "aptitude", 
                                        "mathematics", "language", "generalknowledge"))

#  Convert the studentgroup variable into a factor so it can be 
#  used in models as a classification variable.
morel$studentgroup <- as.factor(morel$studentgroup)

#  Load the car library.  You will need to install the car package
#  if you have not already done so.  Use install.packages("car") 

library(car)

#  Use the lm function to fit a one-way ANOVA to each of the
#  response variables

fit.lm <- lm(cbind(aptitude, mathematics, language, 
                   generalknowledge)~studentgroup , data = morel)
summary(fit.lm)

```

```{r}
 # By default, R sets the effect for the first level of  the
 # treatment factor equal to zero.  If we want to do the 
 # same as SAS, we have to use the contrasts option to reset
 # the default to set  the effect for the last level of the 
 # treatment factor to zero.  Here is how to do it.
 
 fit.lmsas <- lm(cbind(aptitude, mathematics, language, 
             generalknowledge)~studentgroup , data = morel, 
             contrasts = list(studentgroup = contr.SAS))
 summary(fit.lmsas)

```


# Compute the F-test for the Wilk criterion
```{r}
fit.manova <- manova(fit.lm)
fit.manova
summary(fit.manova)

```

# Test for univariate normality for each student group
```{r}
apply(morel[morel$studentgroup == 1, -1], 2, shapiro.test)

apply(morel[morel$studentgroup == 2, -1], 2, shapiro.test)

apply(morel[morel$studentgroup == 3, -1], 2, shapiro.test)
```

# Construct Bonferroni confidence intervals
```{r}
#  Compute sample sizes and number of variables

n1 <- dim(morel[morel$studentgroup==1, -1])[1]
n2 <- dim(morel[morel$studentgroup==2, -1])[1]
n3 <- dim(morel[morel$studentgroup==3, -1])[1]

p <- dim(morel[ , -1])[2]
g <- 3

#  Enter the confidence level
level <- 0.95

# Compute Bonferroni Confidence Intervals
# for each student population
m <- p*g*(g-1)/2
level2 <- 1-(1-level)/(2*m)
df <- n1+n2+n3-g 
c_bon  <- qt(level2, df)

# Compute pooled covariance matrix 
# vpool <- fit.manova$SSPE/df ## The variable SSPE is no longer available, we need to use the following way to get the pooled covariance matrix
vpool = summary(fit.manova)$SS$Residuals / df

# Compute Bonferroni limits

xbar1<-sapply(morel[morel$studentgroup==1, -1], mean)
xbar2<-sapply(morel[morel$studentgroup==2, -1], mean)
xbar3<-sapply(morel[morel$studentgroup==3, -1], mean) 


lower_limit_1v2 <- (xbar1-xbar2) - c_bon*sqrt(diag(vpool)*((1/n1)+(1/n2)))
upper_limit_1v2 <- (xbar1-xbar2) + c_bon*sqrt(diag(vpool)*((1/n1)+(1/n2)))
lower_limit_1v3 <- (xbar1-xbar3) - c_bon*sqrt(diag(vpool)*((1/n1)+(1/n3)))
upper_limit_1v3 <- (xbar1-xbar3)  + c_bon*sqrt(diag(vpool)*((1/n1)+(1/n3)))
lower_limit_2v3 <- (xbar2-xbar3) - c_bon*sqrt(diag(vpool)*((1/n2)+(1/n3)))
upper_limit_2v3 <- (xbar2-xbar3)  + c_bon*sqrt(diag(vpool)*((1/n2)+(1/n3)))
 
rbind(lower_limit_1v2, upper_limit_1v2, lower_limit_1v3, upper_limit_1v3,  lower_limit_2v3, upper_limit_2v3)

```