---
title: "Lab3"
author: "Ben Lehmann, Zachary Picchietti"
date: "2024-09-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Question 1

```{r}
library(ggplot2)
library(GGally)
```


```{r}
music <- read.csv("music-plusnew-sub.csv", row.names=1,
stringsAsFactors=FALSE)
```


```{r}
music.sub <- subset(music, Type == "Rock" | Type == "Classical")
```


```{r}
qplot(LAve, data=music.sub, geom="histogram", binwidth=10.0) +
facet_wrap(~Type, ncol=1)

```

a). Classical is skewed to the right while Rock is skewed to the left. They have different center values.
Classical seems to have an outlier while Rock's distribution does not look like it has any outlier.


```{r}
ggplot(music.sub, aes(x = LVar, y = LAve, color = Type)) +
  geom_point(size = 3) +
  labs(title = "Scatterplot of LVar vs LAve",
       x = "LVar",
       y = "LAve") 
```

b). Classical Music have low LVar and high LAve, Rock music is more scattered,
there is some Rock Music with Low LVar and low LAVe, however, higher LVar will have Rock's LAve be around 0


```{r}
music.sub2 <- subset(music, Type == "Rock" | Type == "Classical" |
Type=="New wave")
```



```{r}
ggparcoord(music.sub2, columns=c(4, 3, 5, 6, 7),
groupColumn="Type",
title="Parallel Coordinate Plot: Music Types")

```

c). Classical Music, is higher in LAve and LFreq, but lower in everything else compared to Rock,
    Rock is high in LVar,LFEner. New Wave is lower than rock, we could say Yes that there is evidence.
    
    

Exercise 2

```{r}
fat <- read.csv('bodyfat.csv')
```


```{r}
ggpairs(fat[,1:4])
```

1). The problem is that there is an outlier where an inch is less than 60 and where Weight is more than 300 pounds.


2).
```{r}
filtered_data <- fat[fat$Heightinches > 60 & fat$Weightlbs < 300, ]
ggpairs(filtered_data[,1:4])
```

The Strongest correlation is between Weight and Body Fat, the negative correlation is Inches and Age.
We look at the Graph comparing Wight and body Fat, we see a positive, linear and strong correlation. Years and weight is near 0 for correlation, so there will be not much.

3).

```{r}
ggparcoord(fat, 4:14)
```
One of the outliers is with the variable HeightInInches, you can see that huge dip in the beginning compared to the rest of the graph.


4).
```{r}
ggparcoord(filtered_data, 4:14)
```
We don't see that sharp decline in the beginning with Height because we removed the outlier.The Graph changes, the y-axis interval of values changes and shrinks.



Excercise 3

```{r}
pisamath <- read.csv("pisamathmeans.csv")
pisamath2 <- subset(pisamath, !is.na(acc))
```

```{r}
summary(pisamath)
```

1). 4978 NA in Acc

2).

```{r}
ggpairs(pisamath2, aes(color=Gender))
```


Each variable is strong in terms of Association, they are all above 0.9
The lowest correlation is 0.89 between api and acs, still impressive correlation.
Each Graph shows a strong positive linear correlation